Covariate analysis of CMC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (ACC and DLPFC with CQN normlaisation)

### 2017-04-06 code/Rmd/CMC_ACC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd
Modified data and metadata source for v75 covariate analysis

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn8414424';
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = downloadFile(COUNT_ACC_ID)

@@

# Get merged metadata
METADATA_ID = 'syn2299154' #'syn7574833';#
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = F)

@@

@@

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",

@@

# Filter metadata 
ind = as.character(METADATA[1,] %>% unlist) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',as.character(METADATA[1,] %>% unlist))

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
                LibraryBatch = ACC_RNA_report__Library_Batch,
                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
                MappedReads = ACC_RNA_report__Mapped_Reads, 
                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
                IntronicRate = ACC_RNA_report__Intronic_Rate, 
                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
                GenesDetected = ACC_RNA_report__Genes_Detected,
                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = ACC_RNA_report__rRNA_Rate, 
                TotalReads = ACC_RNA_report__Total_Reads, 
                PercentAligned = ACC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
                GenesDetected = DLPFC_RNA_report__Genes_Detected,
                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
                TotalReads = DLPFC_RNA_report__Total_Reads, 
                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1))

writeLines('Following samples are tagged outliers and removed')
writeLines(paste(c('MSSM_RNA_ACC_BP_24'), collapse = ', '))
METADATA <- METADATA  %>% filter(!(SampleID %in% c('MSSM_RNA_ACC_BP_24')))
```

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.character)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], as.numeric)

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

@@

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

>>>>

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn8414424'
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = downloadFile(COUNT_ACC_ID)

@@

@@

# Get RNA qc metrics from synapse
METADATA_QC_ACC_ID = 'syn2929053'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_ACC_ID
METADATA_QC_ACC = read.table(synGet(METADATA_QC_ACC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  lapply(as.character)

# Get RNA qc metrics from synapse
METADATA_QC_DLPFC_ID = 'syn2279446'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_QC_DLPFC_ID
METADATA_QC_DLPFC = read.table(synGet(METADATA_QC_DLPFC_ID)@filePath, fill = T, header = T, sep = ',') %>%
  lapply(as.character)

METADATA_QC = rbindlist(list(METADATA_QC_ACC, METADATA_QC_DLPFC), use.names = T, fill = T) %>%
  dplyr::filter(is.na(Exclude.)) %>%
  dplyr::select(Sample.RNA.ID, Ribozero.Batch, Library.Batch, Flowcell.Batch, Mapped.Reads, 
                Intragenic.Rate, Exonic.Rate, Intronic.Rate, Intergenic.Rate, Genes.Detected, 
                Transcripts.Detected, Expression.Profiling.Efficiency, rRNA.Rate, Total.Reads, Percent.Aligned) %>%
  dplyr::rename(SampleID = Sample.RNA.ID, RibozeroBatch = Ribozero.Batch, LibraryBatch = Library.Batch, 
                FlowcellBatch = Flowcell.Batch, MappedReads = Mapped.Reads, IntragenicRate = Intragenic.Rate, 
                ExonicRate = Exonic.Rate, IntronicRate = Intronic.Rate, IntergenicRate = Intergenic.Rate, 
                GenesDetected = Genes.Detected, TranscriptsDetected = Transcripts.Detected, 
                ExpProfEfficiency = Expression.Profiling.Efficiency, rRNARate = rRNA.Rate, 
                TotalReads = Total.Reads, PercentAligned = Percent.Aligned)

# Get merged metadata
METADATA_ID = 'syn4484273'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA_ID
METADATA = read.csv(synGet(METADATA_ID)@filePath, fill = T, header = T, check.names = F)

@@

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",

@@

# Filter metadata 
ind = colnames(METADATA) %in% metadataVarOfInterest
METADATA = METADATA[,ind]
colnames(METADATA) = gsub('[^[:alnum:]]','_',colnames(METADATA))

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DNA_genotyping_report__Genotyping_Sample_ID, ACC_RNA_isolation__Exclude_, 
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DNA_genotyping_report__Genotyping_Sample_ID, DLPFC_RNA_isolation__Exclude_, 
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  list(.,METADATA_QC) %>%
  join_all %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

writeLines('Following samples are missing EV information')
writeLines(paste(METADATA$SampleID[is.na(METADATA$EV.1)], collapse = ', '))
METADATA <- METADATA  %>% filter(!is.na(METADATA$EV.1)) 
```

@@

# Convert factor covariates to factors
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)
COVARIATES[,ContCovariates] = lapply(COVARIATES[,ContCovariates], function(x){
  x = as.numeric(as.character(gsub('[\\,\\%]','',x)))
})

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=8, fig.height=8}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
tmp = COVARIATES.CORRELATION$ESTIMATE
tmp[COVARIATES.CORRELATION$PVAL > 0.05] = 0
h = Heatmap(tmp, col = colorRamp2(c(-1,0,1), c('blue','white','red')), name = 'AssocEstimate')
draw(h, heatmap_legend_side = 'left')
```

@@

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples from `r length(unique(COVARIATES$Individual_ID))` unique individuals

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$fi)$E, scale.=T, center = T)

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
